Deciding Bitvector Termination with Program Synthesis

### 2014-06-23 papers/termination/paper.tex

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.  This fragment
consists of $\exists \forall$ formulae where the the universally quantified variables only range over
ground terms and has close ties to the program synthesis problem.  We will make use of this observation
to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
Section~\ref{sec:synthesis}.

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R \cdot \forall x, x' \cdot & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 \cdot \forall x \cdot \exists x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
 \end{align*}
\end{definition}

\begin{definition}[Deterministic Non-Termination Formula]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
 \end{align*} 
\end{definition}
\end{framed}
\end{figure*}

@@

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
described as finding a satisfying assigment to the following formula, which is the general case of
the synthesis fragment:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
\end{definition}

Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
specification function we reduce the problem of proving termination to the problem of
program synthesis.

\subsection{Synthesis Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

$$O(2^{K(f)}$$

>>>>

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

\begin{definition}[Synthesis Fragment]
 A formula is in the synthesis fragment iff it is of the form
 \[
  \exists P, x . \forall y . \sigma(P, x, y)
 \]

 Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
\end{definition}

We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
to employ the program synthesiser described in Section~\ref{sec:synthesis}.

\subsection{Variations of the Halting Problem}

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R . \forall x, x' . & L(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V . \forall x, x' . & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 . \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
 \end{align*}
\end{definition}

\begin{definition}[Deterministic Non-Termination Formula]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
 \end{align*} 
\end{definition}
\end{framed}
\end{figure*}

@@

\subsection{Program Synthesis as Second-Order Satisfaction}

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
described as finding a satisfying assigment to the following formula, which is the general case of
the synthesis fragment:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists P, x . \forall y . \sigma(P, x, y)$$
\end{definition}

Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.

$$O\left(2^{K(f)}\right)$$

\section{Soundness, Completeness and Complexity}
\begin{theorem}
 {\sc Headshot} is sound -- if it says a program terminates then it does.
\end{theorem}

\begin{theorem}
 {\sc Headshot} is complete -- if a program terminates, {\sc Headshot} will find a proof that it does.
\end{theorem}

\begin{theorem}
 {\sc Headshot} is a decision procedure -- it always terminates.
\end{theorem}

\begin{theorem}
 {\sc Kalashnikov} has time complexity $NP^{NP}(K(\sigma))$
\end{theorem}

\begin{theorem}
 {\sc Headshot} uses a single call to {\sc Kalashnikov} and uses an encoding with $O(1)$
 overhead.  It therefore has time complexity $NP^{NP}(K(\tau(P))$ where $\tau(P)$ is the
 termination specification for program $P$.
\end{theorem}

In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

@@
